### 2021-02-25 referee_response.md

> This paper uses data from the Galaxy Zoo project to address the problem of the pitch angle of spiral arms in disc galaxies. It uses citizen science data products of the Galaxy Builder project, in particular those on the pitch angle of spiral galaxies. Interesting results are obtained from data supplied by volunteers, which eventually should be published. However, the paper is not yet very satisfactory, and I recommend that a major revision should be made. I am willing to see the revised version again.

We are pleased the referee thinks our results are interesting and should be published.

> 1. Introduction

> The introduction shows already some problems: in Figure 1 examples are shown of grand design spirals, multi-arm spirals and flocculent spirals. However, it is easy to show that the spirals shown at the right column of Figure 1 are not flocculent, by rotating their image by 180°, and overlaying them on the original with 50% transparency: this shows that the top-right spiral has two arms in the central parts, and is multi-armed in the outer parts, while the bottom-right image is that of a barred spiral with predominantly two arms.

> Another problem concerns the Hubble classification, which for spirals is aptly summarized by Sandage (2005), who writes "It was simplicity itself with wide classification bins that merged seamlessly into one another according to three criteria of size of the central bulge, the degree of resolution into condensations (in the language of Reynolds 1920b, 1927a,b), and the openness of the spiral arms." It is very problematic that the authors think they can replace the criterion of "the degree of resolution into condensations" by "how obvious spiral patterns are". Sandage was consistent in his write-ups, and the same criterion of the degree of resolution can be found in his 1961 "Hubble Atlas of Galaxies". He mentions in his 2005 paper that the Elmegreen & Elmegreen spiral arm classes are "important for the physics, but need not be made general in the [classification] notation".

> OK, this was the 1st paragraph in the Introduction. The second paragraph is hardly better: A first sentence discusses "a majority of the population of young stars ..." but switches subject to spiral arms: they may trigger star formation, though their main role maybe sweeping up material ... The second sentence asserts that "this rearrangement of) disc gas can lead to the formation of disc-like bulges (... pseudobulges ...), which are prevalent in most spiral galaxies (Kruk et al. 2018). However, in the latter paper this rearrangement of gas is attributed to the bar, and not to the spiral arms…

> Let me point out that the motivation for the Lin & Shu (1964) paper was to overcome the 'winding dilemma' posed by differentially rotating spirals such as M31 and NGC 5055 (Prendergast & Burbidge 1960, Oort 1962), and that the statement of preferentially m = 2 armed spirals is in direct contradiction with the numbers given in Figure 1.

> Furthermore, the mechanisms of wave amplification are not the same as those behind the assumption of quasi stationary density wave theory. The term "swing amplification" was coined by Toomre (1981), even though he relied on the Goldreich and Lynden-Bell (1965) work.

> What to make of the sentence "Larger bulges and more massive central black holes have both correlated with more tightly wound spiral arms" (p3, left column, line 47). They don't correlate anymore?

> The next paragraph then jumps to "Section 3.2 examines ...", and "Section 3.3 investigates ...", without explaining what is to come in Section 2 and 3.1…

@@

* 159 total galaxies, 129 of which with identified spiral arms
* 247 total spiral arms
* 238,433 total points
  *  30  galaxies had 0 arm(s)
  *  38  galaxies had 1 arm(s)
  *  68  galaxies had 2 arm(s)
  *  19  galaxies had 3 arm(s)
  *  4   galaxies had 4 arm(s)

This has been entered into the text.

> In Figure 4 I count about 40 1-armed spirals, 22 3-armed spirals, 14 4-armed spirals, and about 63 2-armed spirals (more difficult to count). Why are there so many 1-armed spirals, and why for these, for pitch angles below about 14 degrees, is PHI_arm not equal to PHI_gal ?

 We have found that in some cases, while volunteers drew two arms, the clustering/aggregation code was unable to cluster them, so the final method reported a single arm. This happened in 35 cases (the majority of the 1 arm galaxies), and is a consequence of clustering noisy data. Where arms are detected the pitch angles are reliable, and our method of calculating galaxy averages accounts for this possible bias of error caused by missing arms. However the presence of this fraction of single-identified spiral arm galaxies drives the average number of arms per galaxy down below the real value, and because the most common number of actual arms is two, it is particularly notable in resulting in reports of one-armed galaxies. We have added some text explaining this, and cautioning against over interpreting our reports on the number of arms. We reiterate that this should not significantly impact the results on pitch angle.

@@

This error has been corrected.

> The range of values of PHI_gal is from 8 to about 50 degrees, and there is no indication that the lower limit in the pitch angle is poorly determined. Note that Kennicutt (1981) find a lower limit of 0 degrees. Setting the lower limit to 15 degrees, so as just to be able to conclude that a uniform cot PHI_gal or cot PHI_arm distribution cannot be excluded is thus completely arbitrary. It would be best to display in a second panel the distribution of cot PHI_gal or cot PHI_arm with a lower limit of 8 degrees, so as to alert the reader to the strong influence of this assumption.

 The issue here is that the model itself includes an arbitrary lower pitch angle - Pringle & Dobbs (2019) make no attempt to give any physical motivation for what that minimum angle should be, and in their test against data chose 11.5 degrees as the limit - largely because that's where inspection of the distribution of cot(pitch angle) in the sample they investigate deviates from constant.

 We think the referee makes an excellent point that the previous explanation was a little confusing. What we have really determined in our study here is that the cot(phi) uniform "model" fits the data well when the minimum pitch angle for winding spirals is around 15 degrees. We also determine that if the minimum pitch angle is 10 degrees the model cannot fit our data. This suggests that a minimum pitch angle for this model must be above 10 degrees,(e.g. 15 degrees). That's a specific observation that a more detailed model based on the simple model of Pringle & Dobbs (2019) would have to explain. We have made edits to the text to make this point.

### 2021-02-25 sections/abstract.tex
Minor edits based on Chris L comments

% !TEX root = ../main.tex
Spiral structure is ubiquitous in the Universe, and the pitch angle of arms in spiral galaxies is an important observable in efforts to discriminate between different mechanisms of spiral arm formation and evolution. In this paper, we present a hierarchical Bayesian approach to galaxy pitch angle determination, using spiral arm data obtained through the \textit{Galaxy Builder} citizen science project. We present a new approach to deal with observed large variations in pitch angle between different arms in a single galaxy, which and obtains full posterior distributions on parameters. We make use of our fitted pitch angles to examine previously reported links between bulge and bar strength and pitch angle, finding no correlation in our data. We test a recent model for spiral arm winding, which predicts uniformity of the cotangent of pitch angle between some unknown upper and lower limits, with results supporting this model of transient and recurrent spiral pitch angle.

>>>>

% !TEX root = ../main.tex
Spiral structure is ubiquitous in the Universe, and the pitch angle of arms in spiral galaxies provide an important observable in efforts to discriminate between different mechanisms of spiral arm formation and evolution. In this paper, we present a hierarchical Bayesian approach to galaxy pitch angle determination, using spiral arm data obtained through the \textit{Galaxy Builder} citizen science project. We present a new approach to deal with the large variations in pitch angle between different arms in a single galaxy, which obtains full posterior distributions on parameters. We make use of our pitch angles to examine previously reported links between bulge and bar strength and pitch angle, finding no correlation in our data {\bf (with a caveat that we use observational proxies for both bulge size and bar strength which differ from other work)}. We test a recent model for spiral arm winding, which predicts uniformity of the cotangent of pitch angle between some unknown upper and lower limits, {\bf finding our observations are consistent with} this model of transient and recurrent spiral pitch angle {\bf as long as the pitch angle at which most winding spirals dissipate or disappear is larger than 10$^\circ$.}

### 2021-02-25 sections/method.tex
Minor edits based on Chris L comments

\begin{equation}
  \label{eq:log-spiral}
r = A\,e^{\theta\tan\phi},
\end{equation}
%
where $\phi$ is the arm's pitch angle, $A$ is an amplitude coefficient and $\theta$ is the polar coordinate. \textbf{Different arms in a galaxy could have different values of $\phi$, however for each arm, $\phi$ is assumed to constant with radius}. One method used to obtain a pitch angle of a galaxy is to fit logarithmic spirals to individually identified arm segments and take the weighted mean of their pitch angles (which often vary by upwards of $10^\circ$, \citealt{2014ApJ...790...87D}). Weighting is determined by the length of the arc segment, with longer being assigned higher weights, i.e. for a galaxy where we have identified $N$ arm segments, each with length $L_i$ and pitch angle $\phi_i$

@@

Fourier analysis in one- and two-dimensions (as performed by \citealt{2019arXiv190804246D}, \citealt{2012ApJS..199...33D}, \citealt{2018MNRAS.474.2594M}, \textbf{and dating back to the seminal work of \citealt{ConsidereAthanassoula1988}}) is another widely used method of computationally obtaining galaxy pitch angles. Two-dimensional Fourier methods generally decompose a deprojected image of a galaxy into a superposition of logarithmic spirals between inner and outer annuli \citep{2012ApJS..199...33D} and reports the pitch angle with the highest amplitude as the galaxy's pitch angle. \citet{2020MNRAS.493.3854H} combined Fourier analysis of spiral galaxies with a visual tracing of spiral arms, successfully eliminating observed bias in a sample of toy images of galaxies. It is unclear how the variation between pitch angles of individual arms impacts this measurement. \textbf{We note that while this method is able to model non-logarithmic spirals - as a sum of logarithmic spirals with differing pitch angles, most applications of the model favour picking regions of the galaxy in which the pitch angle is constant with radius - e.g. see Section 4.3.2 of \citealt{2012ApJS..199...33D}.}

>>>>

\begin{equation}
  \label{eq:log-spiral}
r = A\,e^{\theta\tan\phi},
\end{equation}
%
where $\phi$ is the arm's pitch angle, $A$ is an amplitude coefficient and $\theta$ is the polar coordinate. \textbf{Different arms in a galaxy could have different values of $\phi$, however for each arm, $\phi$ is assumed to be constant with radius}. One method used to obtain a pitch angle of a galaxy is to fit logarithmic spirals to individually identified arm segments and take the weighted mean of their pitch angles (which often vary by upwards of $10^\circ$, \citealt{2014ApJ...790...87D}). Weighting is determined by the length of the arc segment, with longer arms being assigned higher weights, i.e. for a galaxy where we have identified $N$ arm segments, each with length $L_i$ and pitch angle $\phi_i$

@@

Fourier analysis in one- and two-dimensions (as performed by \citealt{2019arXiv190804246D}, \citealt{2012ApJS..199...33D}, \citealt{2018MNRAS.474.2594M}, \textbf{and dating back to the seminal work of \citealt{ConsidereAthanassoula1988}}) is another widely used method of computationally obtaining galaxy pitch angles. Two-dimensional Fourier methods generally decompose a deprojected image of a galaxy into a superposition of logarithmic spirals between inner and outer annuli \citep{2012ApJS..199...33D} and reports the pitch angle with the highest amplitude as the galaxy's pitch angle. \citet{2020MNRAS.493.3854H} combined Fourier analysis of spiral galaxies with a visual tracing of spiral arms, successfully eliminating observed bias in a sample of toy images of galaxies. It is unclear how the variation between pitch angles of individual arms impacts this measurement. \textbf{We note that while this method is able to model non-logarithmic spirals -- as a sum of logarithmic spirals with differing pitch angles, most applications use models which assume that the pitch angle is constant with radius, in some cases picking regions of a galaxy in which this is true - e.g. see Section 4.3.2 of \citealt{2012ApJS..199...33D}.}

### 2021-02-25 sections/summary.tex
Minor edits based on Chris L comments

We do not find any link between bar strength and pitch angle \textbf{in our sample. We make use of a novel (not tested) measure of bar strength, so at best this observation} suggests that the primary mechanism driving the evolution of the spirals in our sample is not Manifold theory (see Section \ref{section:morphology-comparison-bar}), \textbf{but cannot completely rule it out.}

Our results are consistent with spiral winding of the form described by \citet{2019arXiv190910291P}, in which spiral arms are transient and recurrent, evolve through mechanisms such as swing-amplification \citep{1965MNRAS.130..125G} and which wind up over time, \textbf{as long as the minimum pitch angle before the majority of arms dissipate or disappear is $\phi_\mathrm{min} = $ {15\degree}}. The assumptions of this model of spiral winding are highly simplistic, and it leaves many unanswered questions: what determines the limits on $\phi$? Is the spiral arm equally apparent at all pitch angles, or is a selection effect present? \textbf{Our observations suggest that any further development of this model needs to predict that the minimum pitch angle, $\phi_\mathrm{min}>${10\degree}.} This result is also not evidence against QSDW, as our distribution of pitch angles may be dictated by other factors such as disk shear.

>>>>

We do not find any link between bar strength and pitch angle \textbf{in our sample. However, rather that a direct measure of bar strength, we make use of an available parameter which correlates with bar strength, so at best this observation} suggests that the primary mechanism driving the evolution of the spirals in our sample is not Manifold theory (see Section \ref{section:morphology-comparison-bar}), \textbf{but cannot completely rule it out.}

Our results are consistent with spiral winding of the form described by \citet{2019arXiv190910291P}, in which spiral arms are transient and recurrent, evolve through mechanisms such as swing-amplification \citep{1965MNRAS.130..125G} and which wind up over time, \textbf{if the minimum pitch angle before the majority of arms dissipate or disappear is $\phi_\mathrm{min} = $ {15\degree}, but rule it out if the minimum pitch angle is $\phi_\mathrm{min} = $ {10\degree}}. The assumptions of this model of spiral winding are highly simplistic, and it leaves many unanswered questions: what determines the limits on $\phi$? Is the spiral arm equally apparent at all pitch angles, or is a selection effect present? \textbf{Our observations suggest that any further development of this model needs to predict that the minimum pitch angle, $\phi_\mathrm{min}>${10\degree}.} This result is also not evidence against QSDW, as our distribution of pitch angles may be dictated by other factors such as disk shear.
